Providing a Unified Software Environment for Canada’s National Advanced Computing Centers

### 2019-02-20 extended-abstract.tex
Latest suggestions from Ryan and Bart

%
% The "title" command has an optional parameter, allowing the author to define a "short title" to be used in page headers.
\title{Providing A Unified Software Environment for Canada’s National Advanced Computing Centers}

@@

In this paper, we present the solution that we created, which has allowed Compute Canada to serve the needs of over 10,000 researchers across the country. It is used on over 20 different clusters with heterogeneous configurations, from processor architectures as old as Intel Nehalem or AMD Opteron all the way up to Intel Skylake, with and without GPUs, with InfiniBand, Ethernet or OmniPath as the network fabric, and with Slurm or Torque/Moab as the scheduler. It presents a unified software environment to the users, providing over 600 different scientific applications that are available in over 4,000 different combinations of version, compiler and CPU architecture.
\end{abstract}

@@

The third principle is that installation procedures should be tracked and reproducible. This means that automation is needed. Nix already addresses this issue; however it mostly targets basic Linux packages. For scientific packages, we also wanted to ensure that the software packages were optimized for specific hardware architectures, which Nix does not. The tool we chose for this task is EasyBuild~\cite{EasyBuild2012,EasyBuild2014,EasyBuild2016}, which we will discuss in Section~\ref{sub:Compiled_software_packages}.

@@

\section{Distribution mechanism}
\label{sec:Distribution_mechanism}
A shared file system such as NFS, Lustre, or GPFS, can provide access to software packages on a single cluster. However, experience in the field of high performance computing shows that parallel file systems are often challenging to operate at scale in a reliable and performant manner, especially for the high I/O operations per seconds caused by applications startup. In addition, a shared file system is a single point of failure for a cluster, which nearly all jobs rely on in order to run. An alternative for a single cluster is to install binary packages on each node, using a mechanism such as RPM. However, this can result in divergent configurations across nodes, may be restricted by available space on the local disk, and requires an additional configuration management layer to keep the packages synchronized.

@@

Snapshots of the stratum 0 are kept for a while, such that it is possible to roll back to a previous version if needed. Files in the repository are compressed, and their hash is computed to allow space saving by deduplicating objects. Hashes also allow CVMFS clients to verify and invalidate their local cache quickly when contents is changed.

The stratum 0 is replicated to multiple strata 1 servers that are the access points of the clusters to the software environment. They provide redundancy and prevent single points of failure. Files are served to the clusters through HTTP and can be cached locally with multiple proxy servers, protecting against an outage of the wide area network. The closest layer of cache can be configured either to use a local disk on the compute nodes, or on a shared file system on in a diskless environment.

\section{Compatibility layer}
\label{sec:Compatibility_layer}
Isolating the software environment from the operating system requires a compatibility layer containing almost everything that an OS would contain. From this layer, we exclude the Linux kernel and its modules, hardware drivers, as well as privileged binaries.

\subsection{Nix}
\label{sub:Nix}
Nix \cite{Nix} is both a package manager and an operating system (NixOS). Reproducibility being one the core principles in Nix, it is designed to depend as little as possible on the underlying operating system. Every software package installed through Nix resides in a path calculated with the hash of the recipe to build it. This ensures very strong reproducibility, but it also makes updating a package in place impossible. While reproducibility is an important consideration in advanced research computing, bugs sometimes need to be fixed, which is difficult due to Nix' design. At the time of writing this paper, we are still using Nix as a compatibility layer, but we are investigating a switch to Gentoo Prefix described next.

\subsection{Gentoo Prefix}
\label{sub:Gentoo_Prefix}
Gentoo Prefix~\cite{Gentoo} is an alternative solution to provide a compatibility layer. It allows installation of most packages that form a typical Linux inside a non-default location. Contrary to Nix however, packages are installed in a standard directory structure. This is helpful when updates are required, and minimizes problems with applications that expect a standard directory structure. At the time of this writing, we are still experimenting with Gentoo Prefix, and it is not used in production.

@@

EasyBuild originated from University of Ghent, but has now over 200 contributors from all around the world. The framework is based on scripted recipes to install packages, usually from source code. It is designed to create optimized binaries, and eases recompilation of multiple versions of a given application using many combinations of compilers or libraries.

@@

\subsection{Handling multiple CPU architectures}
\label{sub:Handling_multiple_CPU_architectures}
Handling multiple CPU architectures is done by installing applications in architecture dependent paths and module trees. We also have a wide selection of software packages that are installed in an architecture-independent fashion. This is the case, for example, for binary packages, packages that are not performance critical, or packages that consists of source files only. The same recipe is typically used for installing the same package for multiple architectures. While each cluster will usually show only a single architecture, users are allowed to select a different architecture if they wish to run exactly the same binary as on another cluster.

\subsection{Packages for scripting languages}
\label{sub:Packages_for_scripting_languages}
For scripting languages that support installing packages, such as Python, R or Perl, we leverage the existing package manager of that language, rather than installing them as modules. In the case of Python, we are building Python wheels~\cite{Wheels}, linking against our software stack. Wheels are binary packages that can be installed in a matter of seconds, similar to CentOS RPM packages. Since they are built by our team, and linked against our optimize software stack, we ensure that they are both easy to installed and offer good performance.

@@

\section{Other benefits}
\label{sec:Other_benefits}
Because of the portability of our software stack and the distribution mechanism, it can be used on any cluster, or any virtual machine on a cloud. For example, for training events, we have been using it to provision small sized replicas of Compute Canada clusters in short-lived virtual clusters deployed on our OpenStack platform. We have also used this portable software environment to expose a cluster environment directly through a browser, via JupyterHub. It can also help support infrastructures owned by various institutions.

\section{Acknowledgments}
\label{sec:Acknowledgments}
We thank the following organizations and individuals for sharing their time and expertise:
\begin{itemize}
	\item All of the Compute Canada team members who make it possible to manage such a large software stack
	\item Robert McLay, from Texas Advanced Computing Center, for always providing quick and efficient support and
debugging for Lmod
	\item Kenneth Hoste from Ghent University, and Åke Sandgren from Umeå University, for many discussions
regarding EasyBuild
    \item All of the EasyBuild maintainers and contributors, without which EasyBuild would not have nearly as many supported recipes.
	\item Kuang Chung Chen and Bryan Caron, from McGill University, for initiating the groundwork that lead to this
project
	\item Dave Dykstra, Jakob Blomer, and the CernVM project team for support and development of CVMFS
\end{itemize}

>>>>

%
% The "title" command has an optional parameter, allowing the author to define a "short title" to be used in page headers.
\title{Providing a Unified Software Environment for Canada’s National Advanced Computing Centers}

@@

In this paper, we present the solution that we created, which has allowed Compute Canada to serve the needs of over 10,000 researchers across the country. This solution is used on over 20 different clusters with heterogeneous configurations, on processor architectures ranging from AMD's 2010 Magny-Cours to Intel's 2017 Skylake SP, with or without GPUs, with InfiniBand, Ethernet or OmniPath as the network fabric, and with Slurm or Torque/Moab as the scheduler. This stack provides a unified software environment to users, providing over 600 different scientific applications that are available in over 4,000 different combinations of version, compiler and CPU architecture.
\end{abstract}

@@

The third principle is that installation procedures should be tracked and reproducible. This means that automation is needed. Nix already addresses this issue; however it mostly targets basic Linux packages. For scientific packages, we also wanted to ensure that the software packages were optimized for specific hardware architectures, which Nix can not easily do. The tool we chose for this task is EasyBuild~\cite{EasyBuild2012,EasyBuild2014,EasyBuild2016}, which we will discuss in Section~\ref{sub:Compiled_software_packages}.

@@

\section{Distribution mechanism}
\label{sec:Distribution_mechanism}
A shared file system such as NFS, Lustre, or GPFS, can provide access to software packages on a single cluster. However, experience in the field of high performance computing shows that parallel file systems are often challenging to operate at scale in a reliable and performant manner, especially for the high I/O operations per second caused by applications startup. In addition, a shared file system is a single point of failure for a cluster, which nearly all jobs rely on in order to run. An alternative for a single cluster is to install binary packages on each node, using a mechanism such as RPM. However, this can result in divergent configurations across nodes, may be limited by available space on the local disk, and requires an additional configuration management layer to keep the packages synchronized.

@@

Snapshots of the stratum 0 are kept for a while, such that it is possible to roll back to a previous version if needed. Files in the repository are compressed, and their hash is computed to save space by deduplicating objects. Hashes also allow CVMFS clients to quickly verify and invalidate their local caches when content is changed.

The stratum 0 is replicated to multiple strata 1 servers, which serve content to clients and provide redundancy and load balancing. Files are served to the clusters via HTTP, and each cluster has local caching proxy servers to reduce the load placed on the stratum 1 servers by its worker nodes mitigating the impact of a wide area network outage. The closest layer of cache can be configured to use a local disk on the compute nodes, or a shared file system in a diskless environment.

\section{Compatibility layer}
\label{sec:Compatibility_layer}
Isolating the software environment from the operating system requires a compatibility layer containing almost everything that an OS would provide. From this layer, we exclude the Linux kernel and its modules, hardware drivers, and privileged binaries.

\subsection{Nix}
\label{sub:Nix}
Nix \cite{Nix} is both a package manager and an operating system (NixOS). One of the core principles in Nix is reproducibility, so it is designed to depend as little as possible on the underlying operating system. Every software package installed through Nix resides in a path calculated with the hash of the recipe to build it. This ensures very strong reproducibility, but it also makes updating a package in place impossible. While reproducibility is an important consideration in advanced research computing, bugs sometimes need to be fixed, which is difficult due to Nix' design. This difficulty arises in particular because our Nix-installed packages are linked into software that is not installed via Nix. At the time of writing this paper, we are still using Nix as a compatibility layer, but we are investigating a switch to Gentoo Prefix.

\subsection{Gentoo Prefix}
\label{sub:Gentoo_Prefix}
Gentoo Prefix~\cite{Gentoo} is an alternative solution to provide a compatibility layer. It allows installation of most packages that form a typical Linux operating system in a non-default location. Contrary to Nix however, packages are installed in a standard directory structure. This is helpful when updates are required, and minimizes problems with applications that expect a standard directory structure. At the time of this writing, we are still experimenting with Gentoo Prefix, and it is not used in production.

@@

EasyBuild originated from the University of Ghent, but now has over 200 contributors from around the world. The framework is based on scripted recipes to install packages, usually from source code. It is designed to create optimized binaries, and eases recompilation of multiple versions of a given application using many combinations of compilers or libraries.

@@

\subsection{Handling multiple CPU architectures}
\label{sub:Handling_multiple_CPU_architectures}
Handling multiple CPU architectures is done by installing applications in architecture-dependent paths and module trees. We also have a wide selection of software packages that are installed in an architecture-independent fashion. This is the case, for example, for binary packages, packages that are not performance critical, or packages that consist of source files only. The same recipe is typically used for installing the same package for multiple architectures. While each cluster will usually show only a single architecture, users are allowed to select a different architecture if they wish to run exactly the same binary as on another cluster.

\subsection{Packages for scripting languages}
\label{sub:Packages_for_scripting_languages}
For scripting languages that support installing packages, such as Python, R or Perl, we leverage the existing package manager of that language, rather than installing them as modules. In the case of Python, we are building Python wheels~\cite{Wheels}, linking against our software stack. Wheels are binary packages that can be installed in a matter of seconds, similar to CentOS RPM packages. Since they are built by our team, and linked against our optimized software stack, we ensure that they are easy to install and offer good performance.

@@

\section{Other benefits}
\label{sec:Other_benefits}
Because of the portability of our software stack and the distribution mechanism, it can be used on any cluster, or any virtual machine on a cloud. For example, for training events, we have been using it to provision small replicas of Compute Canada clusters in short-lived virtual clusters deployed on our OpenStack platform. We have also used this portable software environment to expose a cluster environment directly through a browser, via JupyterHub. It can also help support infrastructures owned by various institutions.

\section{Acknowledgments}
\label{sec:Acknowledgments}
We thank the following organizations and individuals for sharing their time and expertise:
\begin{itemize}
	\item All of the Compute Canada team members who make it possible to manage such a large software stack
	\item Robert McLay, from Texas Advanced Computing Center, for always providing quick and efficient support and
debugging for Lmod
	\item Kenneth Hoste from Ghent University, and Åke Sandgren from Umeå University, for many discussions
regarding EasyBuild
    \item All of the EasyBuild maintainers and contributors, without which EasyBuild would not have nearly as many supported recipes.
	\item Kuang Chung Chen and Bryan Caron, from McGill University, for initiating the groundwork in Compute Canada that lead to this project
	\item Dave Dykstra, Jakob Blomer, and the CernVM project team for support and development of CVMFS
\end{itemize}
